---
title: "MIS748 Group Project test"
output: word_document
---

```{r setup, include=FALSE}
setwd('/Users/chrissymo/Documents/MSIS/SPRING 2022/MIS 748/Rfiles')
library(tseries)
library(TSA)
library(tsoutliers)
library("readxl")
library(lmtest)
```

## data transform and identify the possible models

```{r cars}
#gas = read.csv("MONTHLY_GAS_V2.csv")
gas = read_excel("MONTHLY_GAS_V3.xlsx")
#time = as.Date(gas$DATE,tryFormats = "%Y-%m-%d")
x.all = ts(gas$GASREGW,start = c(1991,2),frequency = 12)
date = time(x.all)
#use the data from 1991-02 to 2021-09 for training, and predict 2021-10 to 2022-03 (6 months)
x = ts(gas$GASREGW[1:368],start = c(1991,2),frequency = 12)

par(mfrow = c(1, 3))
plot(x,type="l",main="Gas price trend",xlab = "time",ylab = "price")#non constant mean and variance
acf(x,lag.max = 60)#decays very slow, may have unit root, not stationary
pacf(x,lag.max = 60)
adf.test((x)) #pvalue=0.6781

#log transform
logx = log(x)
plot(logx,type="l",main="log gas price",xlab = "time",ylab = "log(price)")
acf(logx,lag.max = 60)
pacf(logx,lag.max = 60)
adf.test(logx)#p-value = 0.7308,can't reject null, has unit root

#take difference
dlogx = diff(logx)
par(mfrow = c(1, 3))
plot(dlogx,type="l",main="first diff of log gas price")
acf(dlogx,lag.max = 60)#it looks seasonality at 1s(1 year)
pacf(dlogx,lag.max = 60)
adf.test(dlogx) #pvalue=0.01<0.05, reject null, do not have unit root

#take seasonality difference
ddlogx = diff(dlogx,12)
plot(ddlogx,type="l",main="Regular&Seasonality difference of log gas price")
acf(ddlogx,lag.max = 60)
pacf(ddlogx,lag.max = 60)#SAR(3),SMA(2),SMA(4)
#adf.test(ddlogx)
#eacf(ddlogx,13,13)#ARMA(5,5)

###seasonality check #####
par(mfrow = c(1, 1))
boxplot(x~cycle(x), names=c("Jan", "Feb", "Mar", "Apr", "May", "June", "July", "Aug", "Sep", "Oct", "Nov", "Dec"))
###outliers check#######


```

#Seasonality

```{r}
#SAR(3)
out1 = arima(ddlogx,order = c(0,0,0),seasonal = list(order = c(3, 0, 0),period=12))
out1#aic = -990.06
par(mfrow = c(1, 2))
acf(out1$residuals, lag = 60)
pacf(out1$residuals, lag = 60)

#SMA(2)
out2 = arima(ddlogx,order = c(0,0,0),seasonal = list(order = c(0, 0, 2),period=12))
out2#aic = -1037.91
par(mfrow = c(1, 2))
acf(out2$residuals, lag = 60)
pacf(out2$residuals, lag = 60)


#SMA(4)
out3 = arima(ddlogx,order = c(0,0,0),seasonal = list(order = c(0, 0, 4),period=12))
out3#aic = -1038.58
par(mfrow = c(1, 2))
acf(out3$residuals, lag = 60, main="ACF of SMA(4)")#MA(1)
pacf(out3$residuals, lag = 60, main="PACF of SMA(4)")#AR(2)
eacf(out3$residuals)


```

#
#
```{r}
#ARIMA(2,1,0) X (0,1,2)
out4 = arima(logx, order = c(2, 1, 0),seasonal = list(order = c(0, 1, 2),period=12))
out4
#ARIMA(0,1,1) X (0,1,2)-->BEST
out5 = arima(logx, order = c(0, 1, 1),seasonal = list(order = c(0, 1, 2),period=12))
out5
#ARIMA(2,1,0) X (0,1,4)
out6 = arima(logx, order = c(2, 1, 0),seasonal = list(order = c(0, 1, 4),period=12))
out6
#ARIMA(0,1,1) X (0,1,4)
out7 = arima(logx, order = c(0, 1, 1),seasonal = list(order = c(0, 1, 4),period=12))
out7

coeftest(out5)
par(mfrow = c(1, 3))
plot(out5$residuals,type = 'l', main = 'Residuals for ARIMA(0,1,1) X (0,1,2)')
acf(out5$residuals,lag.max = 60, main = 'ACF of Residuals')
pacf(out5$residuals,lag.max = 60, main = 'PACF of Residuals')
Box.test(out5$residuals, type='Ljung',lag=12)#p > 0.05

```

#one step rolling forecast
#ARIMA(0,1,1) X (0,1,2)-->better
```{r pressure, echo=FALSE}
#one step ahead rolling forecast
source("rolling.forecast.R")
#ARIMA(2,1,0) X (0,1,2)
rolling.forecast(logx,1,length(logx)-100,c(2, 1, 0),seasonal = list(order=c(0,1,2),period=12))

##ARIMA(0,1,1) X (0,1,2)
rolling.forecast(logx,1,length(logx)-100,c(0, 1, 1),seasonal = list(order=c(0,1,2),period=12))

#ARIMA(2,1,0) X (0,1,4)
rolling.forecast(logx,1,length(logx)-100,c(2, 1, 0),seasonal = list(order=c(0,1,4),period=12))

##ARIMA(0,1,1) X (0,1,4)
rolling.forecast(logx,1,length(logx)-100,c(0, 1, 1),seasonal = list(order=c(0,1,4),period=12))
```

##regular prediction
```{r}
pp = predict(out5, 6)
pred = exp(pp$pred)   #prediction
pred.upp = exp(pp$pred+2*pp$se)  #upper bound for prediction
pred.low = exp(pp$pred-2*pp$se)  #lower bound for prediction
rr = range(c(gas$GASREGW, pred, pred.upp, pred.low))	#find the minimum and maximum y values in your plot
#pred.time = c(as.Date("2022-04-04"),as.Date("2022-04-11"),as.Date("2022-04-18"),as.Date("2022-04-25"))

par(mfrow = c(1, 1))
##plot original data 
plot(x.all,type="l",main="Gas price prediction",xlab = "time",ylab = "price",ylim = rr)
points(x.all,pch=3)
##add prediction
lines(pred,col="blue")
points(pred,col="blue",pch=2)
lines(pred.upp, lty=2, col='blue')	#upper bound of predicted interval
lines(pred.low, lty=2, col='blue')	#lower bound of predicted interval
##add fitted value
lines(fitted.values(out5), col="red",lty=2)
points(fitted.values(out5), col="red",pch=2)

legend.text=c("Actual values","Prediction","Fitted values")
legend("topleft", legend.text, lty = c(1,1,2), col = c("black","blue","red"), pch = c(3,2,2))

#axis(1, total.time, format(total.time,"%Y-%m-%d"),cex.axis=.7)
#lines(xxx.time, xxx)	#observed values
#points(pred.time, pred, pch=2, col='red', type='o')	#predicted values


pred
pred.upp
pred.low

```



##intervention analysis using ARIMA(0,1,1) X (0,1,2)
#Financial Crisis of 2007-2009, set Sept 2008 as 1, since the bankruptcy of Lehman Brothers
```{r}
fc.temp = ts(gas$financial_crisis_temp[1:368],start=c(1991,2),frequency = 12)
fc.perm = ts(gas$financial_crisis_perm[1:368],start=c(1991,2),frequency = 12)
n = 368
n1 = 211
#fc.temp = ts(c(rep(0, n1), 1, rep(0, n-n1-1)), start=c(1991,2), frequency=12) #with order(1,0) gradually gone, with order(0,0) a one-time change
#fc.perm=ts(c(rep(0, n1), rep(1, n-n1)), start=c(1991,2), frequency=12) #with order (0, 0), step change with order (0,1) gradually increase then stable


order = c(0,1,1)
seasonal=list(order=c(0,1,2), period=12)

out5.1 = arimax(logx, order=order ,seasonal = seasonal, method="ML")
out5.1
#Permanant change
out7 = arimax(logx, order=order,seasonal = seasonal, xtransf=data.frame(fc.perm), transfer=list(c(0,0)), method="ML")
out7
#Temporary change
out8 = arimax(logx, order=order,seasonal = seasonal, xtransf=data.frame(fc.temp), transfer=list(c(1,0)), method="ML")
out8
#Temporary &Permanant change
out9 = arimax(logx, order=order,seasonal = seasonal, xtransf=data.frame(fc.temp, fc.perm), transfer=list(c(1,0), c(0,0)), method="ML")
out9
aic = c(out5.1$aic,out7$aic,out8$aic,out9$aic)
#
which.min(aic) #4
coeftest(out9)

# Total impact
tc = filter(fc.temp, filter=out9$coef[4],method='recursive',side=1)*out9$coef[5]  #the impact of temporary change, exponential decay
ls = ts(c(rep(0,n1), rep(out9$coef[6],n-n1)), start=c(1991,2), frequency=12)   #the impact of the permanent change

#pdf("impact.pdf", width=8, height=6)
plot(tc+ls, type='o', xlim=c(2008,2010), ylab='impact')

```


#try to do predict using intervention, the performance is not good.
```{r}
#1991-02 to 2009-02, prediction 2009-03 to 2009-09 (6 months)
x.fc = ts(gas$GASREGW[1:217],start=c(1991,2),frequency = 12)
n.fc = length(x.fc)
n.fc1 = 211
x.fc1 = ts(gas$GASREGW[1:n.fc1],start=c(1991,2),frequency = 12)
out.fc1 = arima(log(x.fc1),order=c(0,1,1),seasonal=list(order=c(0,1,2), period=12))
#box test reject null, so increase order
out.fc1 = arima(log(x.fc1),order=c(0,1,2),seasonal=list(order=c(0,1,2), period=12))

pulse = ts(c(rep(0, n.fc1), 1, rep(0, n.fc-n.fc1-1)), start=c(1991,2), frequency=12) #with order(1,0) gradually gone, with order(0,0) a one-time change
step=ts(c(rep(0, n.fc1), rep(1, n.fc-n.fc1)), start=c(1991,2), frequency=12) #with order (0, 0), step change with order (0,1) gradually increase then stable

order=c(0,1,1)
seasonal=list(order=c(0,1,2), period=12)


m00= arimax(log(x.fc), order=order, seasonal=seasonal, method="ML")
m00

m01= arimax(log(x.fc), order=order, seasonal=seasonal, xtransf=data.frame(step), transfer=list(c(0,0)), method="ML")
m01
m10= arimax(log(x.fc), order=order, seasonal=seasonal, xtransf=data.frame(pulse), transfer=list(c(1,0)), method="ML")
m10
m11= arimax(log(x.fc), order=order, seasonal=seasonal, xtransf=data.frame(pulse, step), transfer=list(c(1,0), c(0,0)), method="ML")
m11

aic.fc=c(m00$aic, m01$aic, m10$aic, m11$aic)
model=list(m00, m01, m10, m11)
which.min(aic.fc)
model[which.min(aic.fc)]

tc = filter(pulse, filter=m11$coef[4],method='recursive',side=1)*m11$coef[5]  #the impact of temporary change, exponential decay
ls = ts(c(rep(0,n.fc1), rep(m11$coef[6], n.fc-n.fc1)), start=c(1991,2), frequency=12) 

m.pred = arima(log(x.fc), order=order, seasonal=seasonal, xreg=data.frame(tc, step))
m.pred 

h=6

pulse.new = ts(c(rep(0, n.fc1), 1, rep(0, n.fc+h-n.fc1-1)), start=c(1991,2), frequency=12) #expand the impulse function by h
step.new = ts(c(rep(0, n.fc1), rep(1, n.fc+h-n.fc1)), start=c(1991,2), frequency=12) 

tc.new = filter(pulse.new, filter=m11$coef[3],method='recursive',side=1)*m11$coef[4]
newdata=data.frame(tc = tc.new[(n.fc+1):(n.fc+h)], step = step.new[(n.fc+1):(n.fc+h)])

pp=predict(m.pred, h, newxreg=newdata)

time = seq(1991+1/12, 2009+7/12, by=1/12)
time.train = seq(1991+1/12, 2009+1/12, by=1/12)
time.new = seq(2009+2/12, 2009+7/12, by=1/12)
par(mfrow = c(1,1))
plot(time, gas$GASREGW[1:223], xlab='Time', ylab='', xlim=c(1991, 2009), type='o', main='gas prediction after introduce fc intervention')
points(time.train, exp(log(x.fc)-m11$residuals), type='o', pch=2, lty=2, col='red')
abline(v=date[n.fc1+1], lty=2)

points(time.new, exp(pp$pred), type='o', pch=2, lty=2, col='blue')
legend.txt=c("Actual", "Fitted", "Predicted")
legend("topleft", legend.txt, col=c("black", "red", "blue"), pch=c(1,2,3), lty=c(1,2,3))


```




##Covid-19 (Mar 2020)
```{r}

#covid.temp = ts(gas[,5][1:368],start=c(1991,2),frequency = 12)
#covid.perm = ts(gas[,6][1:368],start=c(1991,2),frequency = 12)
order = c(0,1,1)
seasonal=list(order=c(0,1,2), period=12)

n= 368
n2 = 349
covid.temp = ts(c(rep(0, n2), 1, rep(0, n-n2-1)), start=c(1991,2), frequency=12) #with order(1,0) gradually gone, with order(0,0) a one-time change
covid.perm=ts(c(rep(0, n2), rep(1, n-n2)), start=c(1991,2), frequency=12)

#Permanant change
out10 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(covid.perm), transfer=list(c(0,0)), method="ML")
out10

#Temporary change
out11 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(covid.temp), transfer=list(c(1,0)), method="ML")
out11

#Temporary &Permanant change
out12 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(covid.temp, covid.perm), transfer=list(c(1,0), c(0,0)), method="ML")
out12

aic2 = c(out5.1$aic,out10$aic,out11$aic,out12$aic)
# 
which.min(aic2) #4
coeftest(out12)

# Total impact
tc2 = filter(covid.temp, filter=out12$coef[4],method='recursive',side=1)*out12$coef[5]  #the impact of temporary change, exponential decay
ls2 = ts(c(rep(0,n2), rep(out12$coef[6],n-n2)), start=c(1991,2), frequency=12)   #the impact of the permanent change

#pdf("impact.pdf", width=8, height=6)
plot(tc2+ls2, type='o', xlim=c(2020,2021), ylab='impact')

```

#fitted value after introduce covid intervention effect
```{r}

par(mfrow = c(1,1))
plot(date[1:368], x.all[1:368], xlab = 'Time', ylab = '', ylim = rr, type = 'o', main = 'Fitted value after introduce covid effect')
points(time(x), exp(logx-out12$residuals), type = 'o', pch = 2, lty = 2, col = 'red')
abline(v=date[349], lty=2)
legend.txt=c("Actual",  "Fitted")
legend("topleft", legend.txt, col = c("black", "red"), pch = c(1,2), lty = c(1,2))

```


# Prediction without intervention before and after covid
```{r}
n=368
n1=349
h = n-n1

x.covid = ts(logx[1:n1],start = c(1991,2),frequency = 12)

out5.2 = arima(x.covid, order = c(0, 1, 1),seasonal = list(order = c(0, 1, 2),period=12))
pp = predict(out5.2, h)

#date = as.Date(gas[,1], tryFormats="%Y-%m-%d")
date = time(x.all)
pred = exp(pp$pred)
pred.upp = exp(pp$pred+2*pp$se)
pred.low = exp(pp$pred-2*pp$se)
rr = range(c(gas$GASREGW[1:368], pred, pred.upp, pred.low))

par(mfrow = c(1,1))
plot(date[1:368], x.all[1:368], xlab = 'Time', ylab = '', ylim = rr, type = 'o', main = 'Fitted values and predictions before introduce covid effect')
points(time(x.covid), exp(x.covid-out5.2$residuals), type = 'o', pch = 2, lty = 2, col = 'red')
points(time(pred), pred, lty = 1, col = 'blue', type = 'o', pch = 2)
abline(v=date[n1+1], lty=2)
abline(v=date[n1+10], lty=2)
legend.txt=c("Actual", "Predicted", "Fitted")
legend("topleft", legend.txt, col = c("black", "blue", "red"), pch = c(1,2,2), lty = c(1,1,2))
```


##Iraq war, began on 20 March 2003
```{r}
war.temp = ts(gas$iraq_war_temp[1:368],start=c(1991,2),frequency = 12)
war.perm = ts(gas$iraq_war_perm[1:368],start=c(1991,2),frequency = 12)

#Permanant change
out13 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(war.perm), transfer=list(c(0,0)), method="ML")
out13

#Temporary change
out14 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(war.temp), transfer=list(c(1,0)), method="ML")
out14

#Temporary &Permanant change
out15 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(war.temp, war.perm), transfer=list(c(1,0), c(0,0)), method="ML")
out15

aic3 = c(out5.1$aic,out13$aic,out14$aic,out15$aic)
# -1089.469 -1087.601 -1086.281 -1085.761
which.min(aic3) #1

```

##9/11
```{r}
attack.temp = ts(gas$attack_temp[1:368],start=c(1991,2),frequency = 12)
attack.perm = ts(gas$attack_perm[1:368],start=c(1991,2),frequency = 12)

#Permanant change
out17 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(attack.perm), transfer=list(c(0,0)), method="ML")
out17

#Temporary change
out18 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(attack.temp), transfer=list(c(1,0)), method="ML")
out18

#Temporary &Permanant change
out19 = arimax(logx, order=order, seasonal = seasonal,xtransf=data.frame(attack.temp, attack.perm), transfer=list(c(1,0), c(0,0)), method="ML")
out19

aic4 = c(out5.1$aic,out17$aic,out18$aic,out19$aic)
# 
which.min(aic4) #2
#n= 368
#n4 = 127
# Total impact
tc4 = filter(attack.temp, filter=out18$coef[4],method='recursive',side=1)*out18$coef[5]  #the impact of temporary change, exponential decay
#ls4 = ts(c(rep(0,127), rep(out18$coef[6],(368-127))), start=c(1991,2), frequency=12)   #the impact of the permanent change

plot(tc4, type='o', xlim=c(2001,2003), ylab='impact')

```



#outlier identify
```{r}
x.all = ts(gas$GASREGW,start = c(1991,2),frequency = 12)
out16 = arima(log(x.all), order = c(0, 1, 1),seasonal = list(order = c(0, 1, 2),period=12))

locate.outliers(out16$residuals, pars=coefs2poly(out16))

```



##regression
```{r}
#1991-02-2021-11
cpi = read.csv("cpi.csv")
oil = read.csv("oil.csv")
#sp = read.csv("sp500.csv")

#full dataset is 1991-02 to 2021-11, since crude oil do not have data after 2021-11
rgas = ts(gas$GASREGW[1:370],start=c(1991,2),frequency = 12)
rcpi = ts(cpi$CPIAUCSL[530:899],start=c(1991,2),frequency = 12)
roil = ts(oil$POILBREUSDM,start=c(1991,2),frequency = 12)
#rsp = ts(sp[1:370,6],start=c(1991,2),frequency = 12)


par(mfrow=c(3,1))
plot(rgas, main='Monthly Gas price', xlab='', ylab='')
plot(rcpi, main='Monthly CPI', xlab='', ylab='')
plot(roil, main='Monthly crude oil price', xlab='', ylab='')
#plot(rsp, main="Monthly S&P 500")

pairs(cbind(Gas = rgas,CPI = rcpi, CrudeOil = roil))

#n = length(rgas)
#data = cbind(rgas, rcpi, roil)
data = cbind(rgas, log(rcpi), log(roil))

train = data[1:365, ]
new.data = data[366:370, 2:3]
lm = lm(rgas~., data = as.data.frame(train))
summary(lm)
e = lm$residuals

plot(e, type='l')

par(mfrow = c(1,2))
acf(e,lag=60)
pacf(e,lag=60)

##estimate regression and ARMA parameters together
rgas1 = log(rgas[1:365])
rcpi1 = log(rcpi[1:365])
roil1 = log(roil[1:365])

out.r0 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(0,0,0))
par(mfrow = c(1,2))
acf(out.r0$residuals)
pacf(out.r0$residuals)
eacf(out.r0$residuals) #ARMA(2,10)

#out.r1 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(2,0,10))
#out.r1 
#coeftest(out.r1)
#out.r2 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(2,0,10),fixed = c(NA,NA,0,0,0,NA,0,0,0,0,0,0,NA,NA,NA))
#out.r2 
#out.r3 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(2,0,1))
#out.r3 

out.r4 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(0,1,0),seasonal=list(order=c(0,1,0), period=12))
acf(out.r4$residuals,lag.max = 60)
pacf(out.r4$residuals,lag.max = 60)#SAR(2),SMA(1)
out.r5 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(0,1,0),seasonal=list(order=c(0,1,1), period=12))#SMA(1) is better
acf(out.r5$residuals,lag.max = 60)
pacf(out.r5$residuals,lag.max = 60)
eacf(out.r5$residuals)#ARMA(0,1)
out.r6 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(0,1,1),seasonal=list(order=c(0,1,1), period=12))

coeftest(out.r6)
#out.r7 = arima(rgas1, xreg=cbind(rcpi1, roil1), order=c(1,0,0),seasonal=list(order=c(0,1,2), period=12))

par(mfrow = c(1,2))
acf(out.r6$residuals,lag=60)
pacf(out.r6$residuals,lag=60)
#dev.off()
Box.test(out.r6$residuals,type="Ljung",lag=12)

```


```{r}
pp = predict(out.r6,5, newxreg=data.frame(new.data))		#prediction for e with time series model
pred = exp(pp$pred)
pred.upp = exp(pp$pred+2*pp$se)
pred.low = exp(pp$pred-2*pp$se)
pp1=predict(lm, as.data.frame(new.data), interval="prediction")		

nn = length(rgas1)
nt = 5	#forecast horizon
nb = 30	#number of data points you want to plot
tt = (nn-nb):nn	#indexes of data points you want to plot
xxx = rgas[tt]		#data you want to plot
rr = range(c(xxx, pred, pred.upp, pred.low, pp1))	#find the minimum and maximum y values in your plot

#pdf('pred.pdf',width=8,height=4)
par(mfrow = c(1,1))
plot(tt, xxx, pch=1, xlim=c(nn-nb, nn+nt), ylim=rr, main='Gas price prediction using ts regresion', ylab='Gas price', xlab='Time')	
lines(tt, xxx)	#observed values
points(366:370, pred, pch=2, col='blue', type='o')	#ts predicted values
lines(nn+1:nt, pred.low, lty=2, col='blue')	#upper bound of predicted interval
lines(nn+1:nt, pred.upp, lty=2, col='blue')	#lower bound of predicted interval
points(nn+0:nt, rgas[nn+0:nt], type='o')

points(nn+1:nt, pp1[,1], pch=2, col='red', type='o')	#lm predicted values
lines(nn+1:nt, pp1[,2], lty=2, col='red')	#upper bound of predicted interval
lines(nn+1:nt, pp1[,3], lty=2, col='red')	#lower bound of predicted interval

#points(nn+0:nt, rgas[nn+0:nt], type='o')

legend.text=c("Actual value", "Prediction-lm", "Prediction-ts")
legend("topleft", legend.text, col=c("black", "red", "blue"), lty = rep(1, 3), pch = c(1,2,2))


```

